---
title: "Module 5 Instructor Notes"
author: "Carolyn Koehn"
format: pdf
---

# Data

## `ID_employment.csv`

These data were retrieved using `tidycensus`.

```{r}
#| eval: false

library(tidycensus)

emply <- get_acs(geography = "county",
                    state = "ID",
                    year = 2017,
                    survey = "acs5",
                    table = "C24030")

write.csv(emply, "data/ID_employment.csv")
```

## `ID_employment_codes.csv`

These data were retrieved from `tidycensus` and simplified.

```{r}
#| eval: false

library(tidyverse)

emply_codes <- load_variables(2017, "acs5") %>%
  filter(str_detect(name, "C24030")) %>%
  select(c(name, label)) %>%
  rename("variable" = "name",
         "variable_meaning" = "label") %>%
  mutate(variable_meaning = str_remove(variable_meaning, "Estimate!!"))

write.csv(emply_codes, "data/ID_employment_codes.csv", row.names = FALSE)
```

## `ID_county-to-county-2013-2017-ins-outs-nets-gross.csv`

These data were downloaded from https://www.census.gov/data/tables/2017/demo/geographic-mobility/county-to-county-migration-2013-2017.html. The dataset is titled "In-, Out-, Net, and Gross Migration [33.5 MB]" under the heading "County-to-County Migration Flows.

After downloading, I saved the Idaho tab as a CSV file.

# Teaching notes

- I had students add the data for their counties to a shared Google sheet so that students could compare their counties with others. You may wish to record data in a different way. If you do, you will want to find and replace mentions of the shared Google sheet throughout these modules.
- I had students email me their finished PDFs. If you have a different way of collecting assignments, you will want to change the instructions at the end of the module to reflect that.